---
title: "Usage of ZygosityPredictor"
author: "Marco Rheinnecker"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{ZygosityPredictor_Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The software package ZygosityPredictor allows to predict how many copies of a 
gene are affected by mutations, in particular small variants (single nucleotide 
variants, SNVs, and small insertions and deletions, Indels). In addition to the 
basic calculations of the affected copy number of a variant, ZygosityPredictor 
can phase multiple variants in a gene and ultimately make a prediction if and 
how many wild-type copies of the gene are left. This information proves to be 
of particular use in the context of translational medicine. For example, in 
cancer genomes, ZygosityPredictor can address whether unmutated copies of 
tumor-suppressor genes are present. ZygosityPredictor was developed to handle 
somatic and germline small variants. In addition to the small variant context, 
it can assess larger deletions, which may cause losses of exons or whole genes.

# Important

To use ZygosityPredictor, the input data has to meet some requirements. The 
most important one is that all input variants, SNVs, Indels and sCNAs must 
originate from a fully clonal tumor. If indications are present that a sample
consists of several sub clones (e.g. in tumor samples), the results must be seen 
with caution as the implementation is founded on the assumption of clonality.
The second important requirement applies mainly to the second part of the tool
and will thereofre be explained in section "Predict Zygosity"

# Installation

The following code can be used to install ZygosityPredictor. The installation
needs to be done once. 

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("ZygosityPredictor")
```


# load example data

To demonstrate the use of ZygosityPredictor, NGS data from the Seq-C2 project 
were used [1]. In the following chunk, all required datalayer of the WGS_LL_T_1 
sample are loaded. The variants are loaded as GRanges objects, one for somatic 
copy number alterations (GR_SCNA), one for germline- and one for somatic small 
variants (GR_GERM_SMALL_VARS and GR_SOM_SMALL_VARS). The input formats will be 
discussed in more detail in section 4. 

```{r}
library(ZygosityPredictor)
library(dplyr)
library(stringr)
library(GenomicRanges)

# file to sequence alignment 
FILE_BAM <- system.file("extdata", "ZP_example.bam", 
                        package = "ZygosityPredictor")
VCF <- system.file("extdata", "ZP_example_chr7.vcf.gz", 
                        package = "ZygosityPredictor")
# meta information of the sample
PURITY <- 0.98
PLOIDY <- 1.57
SEX <- "female"
# variants
data("GR_SCNA")
data("GR_GERM_SMALL_VARS")
data("GR_SOM_SMALL_VARS")
# used gene model
data("GR_GENE_MODEL")
data("GR_HAPLOBLOCKS")

```

# Calculation of affected copies of a variant

Two functions are provided to calculate how many copies are affected by single 
small variants, based on two formulas, one for germline variants and one for 
somatic variants.

## Germline variants

To calculate the affected copies for a germline variant by using 
```aff_germ_copies()```, the following inputs are required:

* **af**: numeric; between 0 and 1; calculated allele frequency of the variant 
in the tumor sample
* **tcn**: numeric; total copy number at the position of the variant
* **purity**: numeric; between 0 and 1; purity or tumor cell content of the 
tumor sample
* **c_normal**: numeric; expected copy number at the position of the variant 
in normal tissue, 1 for gonosomes in male samples, and 2 for male autosomes and 
all chromosomes in female samples. (The function can also assess the c_normal 
parameter by itself, but then the following two inputs must be provided: 
chr and sex)
* **chr**: (only if c_normal is not provided) character; can be either a single 
number or in the “chr1” format; chromosome of the variant
* **sex**: (only if c_normal is not provided) character; either “male” or 
“female” / “m” or “f”; sex of the sample
* **af_normal**: (default 0.5) numeric; allele-frequency of germline variant in 
normal tissue. 0.5 represents heterozygous variants in diploid genome, 1 would
be homozygous. Could be relevant if germline CNVs are present at the position.
Then also the c_normal parameter would have to be adjusted.


the output is a numeric value that indicates the affected copies.

```{r}
## as an example we take the first variant of our prepared input data and 
## extract the required information from different input data layer
## the allele frequency and the chromosome can be taken from the GRanges object

AF = elementMetadata(GR_GERM_SMALL_VARS[1])[["af"]]
CHR = seqnames(GR_GERM_SMALL_VARS[1]) %>%
  as.character()

## the total copy number (tcn) can be extracted from the CNV object by selecting
## the CNV from the position of the variant

TCN = elementMetadata(
  subsetByOverlaps(GR_SCNA, GR_GERM_SMALL_VARS[1])
  )[["tcn"]]

## purity and sex can be taken from the global variables of the sample
## with this function call the affected copies are calculated for the variant
aff_germ_copies(af=AF,
                tcn=TCN,
                purity=PURITY,
                chr=CHR,
                sex=SEX)
```

## Somatic variants

To calculate how many copies are affected by a somatic variant by 
```aff_som_copies()```, the same inputs are required, but a different formula 
is evaluated:
```{r}
## the function for somatic variants works the same way as the germline function

AF = elementMetadata(GR_SOM_SMALL_VARS[1])[["af"]]
CHR = seqnames(GR_SOM_SMALL_VARS[1]) %>%
  as.character()
TCN = elementMetadata(
  subsetByOverlaps(GR_SCNA, GR_SOM_SMALL_VARS[1])
  )[["tcn"]]

aff_som_copies(af=AF,
               chr=CHR,
               tcn=TCN,
               purity=PURITY,
               sex=SEX)
```



## Calculate affected copies of a set of variants

In order to apply the previously mentioned functions to a whole set of variants 
and calculate the affected copies, the following code can be used.
```{r}
## as an example we calculate the affected copies for the somatic variants:
GR_SOM_SMALL_VARS %>%
  ## cnv information for every variant is required.. merge with cnv object
  IRanges::mergeByOverlaps(GR_SCNA) %>% 
  as_tibble() %>%
  ## select relevant columns
  select(chr=1, pos=2, gene, af, tcn) %>%
  mutate_at(.vars=c("tcn", "af"), .funs=as.numeric) %>%
  rowwise() %>%
  mutate(
    aff_copies = aff_som_copies(chr, af, tcn, PURITY, SEX),
    wt_copies = tcn-aff_copies
  )
```

There s also the function ```predict_per_variant``` that basically does the same
with slightly adjusted inputs. 

```{r}
predict_per_variant(purity=PURITY, 
                    sex=SEX,
                    somCna=GR_SCNA,
                    somSmallVars=GR_SOM_SMALL_VARS)

```




# Predict Zygosity

In this section, we will use the WGS_LL_T_1 dataset from the Seq-C2 project as 
an example to investigate whether mutations in the following genes result in 
total absence of wildtype copies. The genes which were selected as an example 
for the analysis are shown below. The example data set was reduced to these 
genes.

* TP53
* BRCA1
* TRANK1
* TRIM3
* JUP
* CDYL
* SCRIB
* ELP2

## Format of input data

Some inputs are optional, while others are compulsory. The latter are labeled 
with “**”. Of note, ZygosityPredictor is applied downstream of variant calling, 
therefore the variant calls, including information on identified somatic copy 
number aberrations (sCNAs), have to be provided. The inputs can be divided into 
five classes:


* File paths:
  + **bamDna\*\* **: character; path to indexed alignment (.bam format)
  + **bamRna**: character; path to rna-sequencing data (.bam format). 
  + **vcf**: character, or character vector containing several vcf file paths; 
  path to variant call file (.vcf.gz format) or .vcf. If provided also input
  **refGen** is required.
  for extended SNP phasing if variants on the same gene are too far away from
  each other for direct phasing
* Sample meta information:
  + **purity\*\* **: numeric; between 0 and 1; indicates purity or tumor cell 
  content of the sample
  + **ploidy**: numeric; ground ploidy of the sample
  + **sex\*\* **: character; “male” or “female” / “m” or “f”; sex of the 
  patient the sample was taken from
* Variants
  + **somCna\*\* **: GRanges object; containing all genomic segments (sCNA) 
  with annotated total copy number (default metadata column name *“tcn”*, 
  custom name can be provided by COLNAME_TCN) and information about LOH 
  (default column name *“cna_type”*, custom name can be provided by 
  COLNAME_CNA_TYPE). The cna_type column should contain the string “LOH” if 
  loss-of heterozygosity is present at the segment. If large deletions should 
  be included to the analysis the total copy number has to be decreased 
  accordingly. If the total copy number is smaller than 0.5, the tool will 
  assume a homozygous deletion. An incomplete deletion is assumed if at least 
  one copy is lost compared to the ploidy of the sample (works only if the
  ploidy is provided as an input)
  + **somSmallVars**: GRanges object; containing all somatic small variants. 
  Required metadata columns are: reference base (*“REF”/”ref”*), alternative 
  base (*“ALT”/”alt”*), allele frequency in the tumor sample (raw allele 
  frequency, i.e. as measured in the tumor sample; not the corrected allele 
  frequency in the supposedly pure tumor) (*“AF”/”af”*), gene (*“gene”/”GENE”*, 
  according to the used gene model (GENCODE39 in the example data) and the 
  annotation provided below). If no relevant somatic small variants are 
  present, can also be NULL or not provided.
  + **germSmallVars**: GRanges object; Analogous to GR_SOM_SMALL_VARS. If no 
  relevant germline small variants are present, can also be NULL or not 
  provided.
* Used Gene model
  + **geneModel\*\* **: GRanges object; containing the gene model for the used 
  reference genome. Required metadata columns are: *"gene"*. Artificially 
  restricting the gene model can be used to tell the tool which genes to 
  analyze. In the case of this vignette, the object contains only the genes we 
  selected.
* Haploblocks
  + **haploBlocks**: GRanges object; containing predefined haplock region, i.e.
  genomic regions in which SNPs could be phased to one specific haplotype. By 
  providing this haploblock together with the SNPs and its genotype annotation
  (1|0 or 0|1) via input **vcf**, haploblock phasing will be enabled.
* Options
  + **logDir**: character; If provided, detailed output will be stored in it
  + **includeIncompleteDel**: logical, default=TRUE; Should incomplete 
  deletions (monoallelic deletions in a diploid sample) be included in the 
  evaluation? Since these often span large parts of the chromosomes and will 
  lead to many affected genes, it can be advisable to include or exclude them, 
  depending on the research question.
  + **includeHomoDel**: logical, default=TRUE; Should homozygous deletions be 
  included in the evaluation?
  + **showReadDetail**: logical, default=FALSE; If this option is TRUE, another 
  table is added to the output that contains more detailed information about 
  the classification of the read pairs. More detailed information is provided 
  in section 4.3.4.
  + **assumeSomCnaGaps**: logical, default=FALSE; Only required if the somCna 
  object lacks copy number information for genomic segments on which small 
  variants are detected. By default, variants in such regions will be excluded 
  from the analysis as required information about the copy number is missing. 
  These variants will be attached to the final output list in a separate 
  tibble. To include them, this flag must be set TRUE and the ground ploidy 
  must be given as an input. This ground ploidy will then be taken as *tcn* 
  in the missing regions.
  + **byTcn**: logical, default=TRUE; optional if includeHomoDel or 
  includeIncompleteDel is TRUE. If FALSE the tool will not use tcn as a 
  criterion to assign large deletions. It will use the cna_type column and 
  check for indicating strings like HOMDEL/HomoDel/DEL. Some commonly used 
  strings are covered. It is recommended to leave this flag TRUE.
  + **printLog**:  logical, default=TRUE; If TRUE, the tool will print detailed
  information how the assessment is done for each gene.
  + **verbose**: logical, default=FALSE; prints debugging information
  + **colnameTcn**: character; indicating the name of the metadata column 
  containing the tcn information in the somCna object. If not provided 
  the tool tries to detect the column according to default names.
  + **colnameCnaType**: character; The same as for colnameTcn, but for 
  cna-type information.
  + **distCutOff**: numeric, default=5000; if input vcf is provided and SNP
  phasing is performed, this will limt the distance at which the SNP phasing
  should not be tried anymore. As the probability of finding overlapping reads
  at such a long distance is very low and the runtime will increase
  exponentially.
  + **snpQualityCutOff**
  + **refGen**: character, default="hg38"; Required if vcf files is provided. 
  Either "hg38" or "hg19". Relevant for VariantAnnotation vcf loading


## Predict zygosity for a set of genes in a sample

The prediction of zygosity for a set of genes in a sample can be assessed by 
the ```predict_zygosity()``` function.

**Important note**: The runtime of the analysis depends strongly on the number 
of genes to be assessed and on the number of input variants. It is therefore 
recommended to reduce the number of genes to the necessary ones. Also, 
depending on the research question to be addressed, the variants should be 
filtered to the most relevant ones, not only because of runtime considerations, 
but also to sharpen the final result. A large number of mutations in a gene, 
some of which are of little biological relevance or even SNPs, will inevitably 
reduce the validity of the results.

```{r, results = FALSE}
fp <- predict_zygosity(
  purity = PURITY, 
  ploidy = PLOIDY,
  sex = SEX,
  somCna = GR_SCNA, 
  somSmallVars = GR_SOM_SMALL_VARS, 
  germSmallVars = GR_GERM_SMALL_VARS, 
  geneModel = GR_GENE_MODEL,
  bamDna = FILE_BAM,
  vcf=VCF,
  haploBlocks = GR_HAPLOBLOCKS
)
```

## Interpretation of results

Of note, the results displayed here were chosen to explain and exemplify the 
functionality of the tool; biological and medical impact of the specific 
variants had not been a selection criterion.
The result which is returned by the function consists of a list of three 
(or up to six) tibbles. 

* Evaluation per variant
* Evaluation per gene
* Phasing info
* Read pair info (only if showReadDetail=TRUE and logDir is provided)
* Variants not covered by somCna (only if present and no sCNA gap assumption 
was done)
* detailed information about extended SNP phasing if it was performed

The information can also be accesed via the accessor 
functions: ```ZP_ov()``` and ```gene_ov()```. 

```{r}
ZP_ov(fp)
```

The function provides an overview about the evaluations done by 
ZygosityPredictor. One can see that in our case 12 input small variants were 
used to predict 7 gene status. Of those, two got the status all_copies_affected.
By using  ```gene_ov()```, more detailed information can be viewed about how the
tool came to a gene status. What the accessor function does is basically 
filtering the output tibbles to the gene of relevance.

```{r}
gene_ov(fp, OR2J1)
```

The first prompted tibble originates from the eval_per_gene tibble of the output. 
The tibble shown 
below originates from eval_per_variant and shows all input small variants of the
gene. The next tibble originates from main_phasing_info and shows the main
phasing combinations. Finally, the last tibble comes from detailed_phasing_info
and contains every phasing combination including the ones with and between SNPs.

What can be seen is that the final gene status of OR2J1 is wt_copies_left. The
tool predicted around 2.6 remaining wt copies with maximum confidence level of 
1. Below it can be seen that 3 small variants inside the gene were in the input
leading to 3 main phasing combinations. Of those, two were directly phased while 
m2-m3 was phased indirectly via m1-m2 and m1-m3. As all main combinations could 
be solved, the left wt copies could be accurately defined and the gene status 
assigned. In this particular case, the gene status could have been solved even 
without phasing all main combinations by the case we refer to as exclusion.
As can be seen in the main phasing combinations in column: min_poss_wt_cp, none 
of the main phasing combinations could lead to less than 0.5 wt copies even if 
the variants were found on different reads.






### Evaluation per variant

The first result of the function is the evaluation per variant. In this step 
all information required for subsequent steps is annotated and the affected 
copies per variant are calculated. For every variant, the function checks 
whether it already affects all copies of the gene. The format of the output is 
a tibble; the number of rows corresponds to the total 
number of input variants. The tool annotates a few self-explanatory columns 
such as the origin of the respective variant (germline/somatic) or the class 
(snv/ins/del). It also appends information from the sCNA results: the total 
copy number at the position of the variant and the information if a loss of 
heterozygosity is present (cna_type). Also, an ID is assigned to every small 
variant. Then, the genes are numbered consecutively in order to unambiguously 
assign variants to genes in the following analysis. The most important results 
of this step are the calculation of the affected and wildtype copies, as well
as, depending on the data, an initial check of whether a variant already affects 
all copies.
Of note, there can be situations in which left wildtype copies are below 0.5,  
but still this information is not sufficient to predict *“all copies affected”* 
without doubt. Depending on the origin of the variant, further criteria must 
be met (e.g., LOH). The procedure for this first check is shown in the pre_info 
column.


```{r}
# here the new columns are selected and viewed
fp$eval_per_variant %>%
  # these steps are just to have a better overview
  # round numeric columns
  mutate_at(.vars=c("af","tcn", "aff_cp", "wt_cp"),
            .funs = round, 2) %>%
  # to get a better overview, the columns which are already in the inut are 
  # removed
  select(-chr, -pos, -alt, -ref, -af)
```
For example, mutation m1 in *TRANK1* (first line) fulfills all criteria that 
are sufficient for a germline variant to say with a high degree of certainty 
that all copies are affected, which can be seen in the pre_info column. The 
same applies to the somatic *JUP* variant. In the case of the *ELP2*, *CDYL* 
and *SCRIB* genes, there are two or even three variants, neither of which leads 
to complete loss of the wildtype gene, which is why the next step is to check 
whether they together affect all copies. Homzygous deletions (homdels) always 
lead automatically to "all copies affected", whereas incomplete deletions do not 
affect all copies.

### Aggregation per gene

The aggregation at gene level is one of the main features of ZygosityPredictor. 
A prediction of how many copies are affected by mutations is provided for every 
gene. The final prediction is stored in tibble format with three columns: gene, 
status - either “all copies affected” or “wt copies left” - and an info column 
that explains how the prediction was made for that gene.
```{r}
fp$eval_per_gene
```

Each gene appears once in the table. If there are several variants in one gene, 
the info always refers to the one that affects the highest number of copies. If 
most copies are affected by a combination of  two small variants, the info 
refers to both of them. In general, the heterozygous deletions have the lowest 
priority; and if any other small variant or homdel is present in the gene, it 
will always be indicated as the more relevant variant. In the case of *ELP2*, 
*CDYL* and *SCRIB*, several small variants are present, which is why the tool 
attempts to do phasing at the present position. This is indicated by “phasing” 
in the info column. The exact results of the phasing can be viewed in detail 
in the next step.

### Haplotype-phasing results

A haplotype phasing is only performed if more than one variant occurs in a 
gene, each of which does not affect all copies of the gene by itself. The tool 
then tries to phase at this position to check if the variants are located on 
the same or different copies. That means that not at every prediction a phasing 
is done and therefore the following output is not always present.
```{r}
fp$phasing_info
```
In the present case, phasing was performed for three genes separately. 
The output contains information such as the distance between the two variants 
and the tcn at the position. If tcn differs at both positions, for example for 
breakpoint genes, the larger one is displayed here. The status column shows 
whether the mutations were found on the same copy (*same*) or on different 
copies (*diff*). This prediction is determined by whether the mutations are on 
the same reads / read pairs. The column DNA_rds contains the number of 
reads / read pairs that cover both positions. If RNA is available, the number 
of reads / read pairs is also displayed for RNA reads. Depending on whether the 
variants were found on different or the same copies, a final prediction for the 
remaining wildtype-copies can be calculated from the affected copies of both 
variants. 
This final calculation can be found in the column *left_wt_cp*. Only if the 
variants are found on different copies and the remaining wildtype-copies are 
below 
0.5, it can be assumed that no wildtype copies are left. 
The four columns *both*, 
*none*, *mut1*, and *mut2* indicate how many of the reads / read pairs belong 
to each category: Both variants detected, none, only the first, or only the 
second. RNA reads that do not contain an exon necessary for phasing, e.g., due 
to alternative splicing or that map somewhere entirely different in the genome, 
end up in a category called *non-overlapping*. The category *dev_var* contains 
reads / read pairs that carry a different variant than the one to be expected 
at the position. The output for *ELP2* shows how genes are handled in which 
more than 2 variants occur. Each mutation is then combined once with each other 
to perform the phasing on each combination. In the case of *ELP2* three 
variants are present, resulting in three possible combinations. Two of the 
variants are located close to each other (*m1* and *m2*), while the third one 
has a distance of almost 9000 basepairs to them. Therefore, it is only possible 
to assign a status to the combination of the close variants. The two other 
combinations get the status "null" as no overlapping reads / read pairs are 
present. Theoretically, the tool can attempt phasing if no reads overlap 
both positions. For such distant variants, intermediate SNPs can be used, 
which then have to be provided by the input *vcf*.

### Detailed info about used read pairs

This output is only provided if the option showReadDetail is set TRUE. It 
consists of a table containing every read pair that was used to perform 
haplotype phasing. In more detail, the read name of the read in the bam file is 
provided with the classification in one of the four categories mentioned before 
(*both, none, mut1, mut2, no_overlap, dev_var*).

# References

1. Fang LT, Zhu B, Zhao Y, Chen W, Yang Z, Kerrigan L, Langenbach K, de Mars M, 
Lu C, Idler K, et al. Establishing community reference samples, data and call 
sets for benchmarking cancer mutation detection using whole-genome sequencing. 
Nature Biotechnology. 2021;39(9):1151-1160 / PMID:34504347
2. Lawrence M, Huber W, Pages H, Aboyoun P, Carlson M, et al. (2013) Software 
for Computing and Annotating Genomic Ranges. PLoS Comput Biol 9(8): e1003118. 
doi:10.1371/journal.pcbi.1003118*
3. Wickham H, François R, Henry L, Müller K (2022). _dplyr: A Grammar of Data 
Manipulation_. R package version 1.0.10, 
<https://CRAN.R-project.org/package=dplyr>.
4. Wickham H (2022). _stringr: Simple, Consistent Wrappers for Common String 
Operations_. 
https://stringr.tidyverse.org, https://github.com/tidyverse/stringr.

```{r}
sessionInfo()
```

